---
title: "Randomized testing with Quickcheck"
author: "Revolution Analytics"
output:
  ioslides_presentation:
    widescreen: true
    incremental: no
    transition: 0.01
    keep_md: yes
    logo: RevolutionAnalytics.png
---

```{r, echo=FALSE}
library(knitr)
read_chunk("../pkg/tests/tutorial.R")
opts_chunk$set(echo=TRUE, tidy=FALSE, comment="", cache=FALSE, error=FALSE)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
		x = unlist(stringr::str_split(x, '\n'))
		if (length(x) > n) {
			# truncate the output
			x = c(head(x, n), '....\n')
		}
		x = paste(x, collapse = '\n') # paste first n lines together
	}
	hook_output(x, options)
})
opts_chunk$set(out.lines = 5)
#options(width = 80)
```



# Testing
 
## {.build}

```{r library-testthat}
```

```{r test_that, eval=FALSE}
```

```{r for-test_that}
```

# Randomized Testing

## {.build}

```{r test, eval=FALSE}
```

```{r test, echo=FALSE}
```

# Advantages

## {.build}

```{r sample.size}
```

# Assertions

##  {.build}

```{r, eval=FALSE}
function(x = rdouble(), y = rdouble()) all(x + y == y + x))
```

```{r, eval=FALSE}
function(x = rlist()) identical(x, rev(rev(x)))
```

```{r, eval=FALSE}
forall(x = rdouble(), y = rdouble(), all(x + y == y + x))
```

```{r, eval=FALSE}
forall(x = rcharacter(), stop(x))
```


```{r expect}
```

# Failure

## {.build}

```{r output, error = TRUE}
```

## {.build}

```{r return-value}
```

## {.build}

```{r repro}
```


# Writing Tests

## {.build}

```{r, results='hide'}
test(forall(x = rinteger(), identical(x,x)), about = "identical") 
```

```{r, results='hide'}
test(forall(x = rany(), identical(x,x)), about = "identical")
```


# Random Data Generators

## {.build}

```{r rdouble}
```

## {.build}

```{r rdouble-2}
```

```{r rdouble-mean-sd}
```

## {.build}

```{r rinteger-min-max}
```

```{r rinteger-unnamed}
```


```{r rinteger-matching}
```

## {.build}


```{r rdouble-runif}
```

```{r rdouble-formula}
```

## {.build}

```{r rdouble-Curry}
```

## {.build}

```{r rdouble-size-max}
```


```{r rdouble-size-min-max}
```


```{r rdouble-size-function}
```


```{r rdouble-size-formula}
```

## {.build}

```{r rdouble-size-constant}
```


```{r rdouble-size-formula-2}
```

<!--
## {.build}

```{r rinteger-formula}
```
-->


## {.build}

|function| description|
|---|---|
`rany` | any R object|
`ratomic` | any atomic vector|
`rcharacter` | character
`rdata.frame` |data.frame|
`rDate` |Date|
`rdouble` |double|
`rfactor` |factor|

## {.build}

|function| description|
|---|---|
`rinteger` | integer|
`rlist` | list|
`rlogical` | logical
`rmatrix` | matrix
`rnamed` | random names for a vector
`rnumeric` | any numeric|
`rraw` |raw
`rsample` | sample from vector

## {.build}

|function| description|
|---|---|
`constant` | trivial constant RDG
`mixture` | mixture RDG
`named` | named version of any RDG


<!--

# Custom Generators

## {.build} 

```{r is.reciprocal.self.inverse}
```


```{r test-is.reciprocal.self.inverse}
```

## {.build}

```{r corner-cases}
```

## {.build}

```{r rdoublex}
```

## {.build}

```{r test-rdoublex}
```

## {.build}
```{r, eval = FALSE}
rlist(rdouble)
```
```{r, eval = FALSE}
rlist(function() rlist(rdouble))
```

```{r, eval = FALSE}
rlist(~rlist(rdouble))
```

-->

## Repo

- http://github.com/RevolutionAnalytics/quickcheck
