---
title: "Assertion-based testing with Quickcheck"
output: 
  html_document:
    keep_md: true
---

`r library(knitr)`
`r opts_chunk$set(echo=TRUE, tidy=FALSE, comment="", cache=TRUE, autodep=TRUE, error=TRUE)`


## Introduction

Quickcheck was originally a package for the language Haskell aiming to simplify the writing of tests. The main idea is the automatic generation of tests based on assertions a function needs to satisfy and the signature of that function. The idea spread to other languages and is now implemented in R with this package. Because of the differences in type systems between Haskell and other languages, the original idea morphed into something different for each language it was translated into, and R is no different. The main ideas retained are that tests are based on assertions and that the developer should not have to specify the specific inputs and output values of a test. The difference from Haskell is that the user needs to specify the type of each argument to a function with the optional possibility to fully specify its distribution. The main function in the package, `unit.test`, will randomly generate input values, execute the function to be tested and verify an assertion about the input output relation. The advantages are multiple:

  - each test can be run multiple times on different data points, improving coverage and ability to detect bugs;
  - test can run on large size inputs, possible but impractical in non-randomized testing;
  - assertions are more self-documenting that specific examples of the I/O relation. In fact, enough assertions can constitute a specification for the function being tested (that's why Quickcheck is sometimes said to implement specification-based testing, which I think is setting a discouragingly high bar for no reason).
  
## First example

Let's start with something very simple. Let's say we just wrote the function identity. Using the widely used testing package `testhat`, one would like to write a test like:

```{r, eval=FALSE}
library(testthat)
test_that("identity test", expect_identical(identity(x), x))
```

That in general doesn't work because `x` is not defined. What was intended was something like a quantifier *for all legal values of `x`*, but there isn't any easy way of doing that. So a developer has to enter some values for `x`.

```{r}
for(x in c(1L, 1, list(1), NULL, factor(1)))
  test_that("identity test", expect_identical(identity(x),x))
```

But there is no specific reason to pick those specific examples, testing on more data  points or larger values would increase the clutter factor, a developer may inadvertently inject unwritten assumptions in the choice of data points etc. `quickcheck` can solve or at lease alleviate all those problems

```{r}
library(quickcheck)
unit.test(assertion = function(x) identical(identity(x), x), generators = list(rinteger))
```

We have supplied an assertion, that is a function returning a length-1 logical vector, where `TRUE` means *passed* and `FALSE` means *failed*, and a list of generators, one for each argument of the assertion -- use named or positional arguments as preferred.
What this means is that we have tested `identity` for this assertion on random integer vectors. We don't have to write them down one by one and later we will see how we can affect the distribution of such vectors, to make them potentially huge in size and contained values. We can also repeat the test multiple times on different values with the least amount of effort, in fact, we have already repeated the test 10 times, which is the default. But if 100 times is required, no problem:

```{r}
unit.test(assertion = function(x) identical(identity(x), x), generators = list(rinteger), sample.size = 100)
```

Done! You see, if you had to write down those 100 integer vectors one by one, you'd never have time to. But, you may object, `identity` is not supposed to work only on integer vectors, why did we test only on those? That was just a starter indeed. Quickcheck contains a whole repertoire of random data generators, including `rinteger`, `rdouble`, `rcharacter` etc for most atomic types, and some also for non-atomic types such as `rlist` and `rdata.frame`. Notable omission is `rfunction` -- don't expect it any time soon. The library is easy to extend with your own generators (for instance, `rnorm` works out of the box) and offers a number of constructors for data generators such as `constant` and `mixture`. In particular, there is a generator `rany` that creates a mixture of all R types (in practice, the ones that `quickcheck` currently knows how to generate, but it's defined in the most general way). That is exactly what we need for our identity test.

```{r}
unit.test(assertion = function(x) identical(identity(x), x), generators = list(rany), sample.size = 100)
```

Now we have more confidence that `identity` works for all types of R objects.

## Ways to define assertions

Unlike `testhat` where you need to construct specially defined *expectations*, `quickcheck` accepts run of the mill logical-valued functions. For example `function(x) x + 0 == x` or `function(x) identical(x, rev(rev(x)))` are valid assertions -- note I said *valid*, albeit they may still fail under certain conditions. If an assertion generates an error, this is also considered a failure. That is `stop` is a valid assertion but always fails. How do I express the fact that this is correct behaviour? Here is where the function `as.predicate` come handy. It allows to convert a `testthat` expectation into a predicate, and expectations, which seem overkill to express the identity of two values,  are a powerful way of expressing less common requirements such as raising exceptions or generating warnings. In the case of `stop`, the appropriate expectations is `expect_error(stop())`. We can use it with `unit.test` as follows:

```
unit.test(
  as.assertion(function(x) expect_error(stop(x))), 
  list(rcharacter))
```

By executing this test successfully we have built confidence that the function `stop` will generate an error whenever called with any `character` argument.


## Ways to define random data generators

## Advanced topics

```{r}
unit.test(function(l) unit.test(function(x,y) {x + y; TRUE}, list(fun(rdouble(size = constant(l))), fun(rdouble(size = constant(l))))), list(rinteger))
```